---
title: "Mutation Calling"
author: "Julie M. Grosse-Sommer and Jarrod D. Hadfield"
output: 
  bookdown::html_document2:
    toc: true
    css: custom.css
bibliography: /Users/jhadfiel/Work/Tex/library/JarLib.bib
---

<script src="hideOutput.js"></script>

<script type="text/javascript">
  // When the document is fully rendered...
  $(document).ready(function() {
    // ...select all header elements...
    $('h1, h2, h3, h4, h5').each(function() {
      // ...and add an id to them corresponding to their 'titles'
      $(this).attr('id', $(this).html());
    });
  });
</script>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if(Sys.info()["nodename"]%in%c("sce-bio-c04553", "sce-bio-c06973", "SCE-BIO-C07467", "S42-76R7VW5J")){
  root<-"/Users/jhadfiel/Work/Ecoli_MA"
}else{
  root<-"/Users/s2221673/Ecoli_MA"
}

# rmarkdown::render(file.path(root, "scripts/Mutation_Calling.Rmd"))
map_reads<-FALSE
call_var<-FALSE
process_vcf<-TRUE
sim_mutations<-FALSE
save<-FALSE
```

In this workbook we detail the workflow for mapping reads and calling, validating and annotating variants in 194 *E. coli* mutation accumulation lines and two replicates of their common ancestor. The raw reads can be found on NCBI's SRA (see [BioProject ID PRJNA1466858](https://www.ncbi.nlm.nih.gov/sra?LinkName=bioproject_sra_all&from_uid=1466858)). A root directory needs to be specified (`~/Work/Ecoli_MA`) and within that directory a subdirectory called raw should exist. The fasta files are to large to be hosted on GitHib but can be downloaded from NCBI's Sequence Read Archive under BioProject ID PRJNA1466858. They should be placed in the raw directory with the fasta files for each library contained within its own subdirectory. The subdirectory should be named after the sample. The sample name is the leading text of the SRA Name field prior to the sequence and the MA line is listed in the SRA Title field (e.g the first SRA is sample 270-94 corresponding to MA Line 26). Details about the samples can be found in meta/sample_submission.csv.

# Mapping Reads

The reads are mapped to the specified reference genome (NCBI RefSeq accession NC\_000913.3, assembly GCF\_000005845.2 (ASM584v2)) using bwa [@Li.2013]. bam files are stored in a subdirectory of root called bam.  If the bam file for a sample exists mapping will be skipped.  The bash script map.sh (in the scripts subdirectory of root) is shown below.

<div class="fold s">
```{r show_map, echo = FALSE, results = 'asis'}

scr <- readLines(file.path(root, "scripts/map.sh"))

cat("```bash\n")
cat(scr, sep = "\n")
cat("\n```")
```
</div>


```{r map, echo = TRUE, eval=map_reads}
system(file.path(root, "scripts/map.sh"))
```

# Calling Mutations

PCR duplicates, secondary reads and non-mapped reads were removed using samtools and variants called using bcftools [@Danecek.2021]. The bash script call.sh (in the scripts subdirectory of root) is shown below. 

<div class="fold s">
```{r show_call, echo = FALSE, results = 'asis'}

scr <- readLines(file.path(root, "scripts/call.sh"))

cat("```bash\n")
cat(scr, sep = "\n")
cat("\n```")
```
</div>


```{r call, echo = TRUE, eval=call_var}
system(file.path(root, "scripts/call.sh"))
```

# Annotating Mutations

We use an R pipeline to filter and annotate variants. Variants at sites with a Phred-scaled score (QUAL) $<20$ are excluded and the remainder are considered possible mutations. These possible mutations and associated meta-data are written to pos\_mutations.csv in a subdirectory of root called meta. Possible mutations with at least 30 reads supporting the alternate allele and less than 5\% of reads supporting the reference allele were considered real mutations without the need for checking. Possible mutations not meeting this strict filter were visualised in the Integrative Genomics Viewer [@Robinson.2011]. The file manual\_check.csv in the subdirectory meta states which of the possible mutations were checked and whether they were validated or not. Using this information a set of real validated mutations are written to real\_mutations.csv in the meta subdirectory.

We can start by organising the necessary objects for annotation

<div class="fold s">
```{r annotate, message=FALSE, warning=FALSE}
library(data.table)
library(VariantAnnotation)  
library(GenomicFeatures) 
library(rtracklayer)      
library(txdbmaker)
library(GenomeInfoDb)

sample_line<-read.csv(file.path(root, "meta/sample_submission.csv"))
# meta-data for samples 

gff_file <- file.path(root, "ref/GCF_000005845.2_ASM584v2_genomic.gff")
fna_file <- file.path(root, "ref/GCF_000005845.2_ASM584v2_genomic.fna")    

# path to gff genome annotation file and FASTA nucleotide sequence for the reference

txdb <- makeTxDbFromGFF(gff_file, format = "gff")
# create database of genomic features

gff <- import(gff_file)
# import the gff file

cds_gr <- cds(txdb,  columns = c("gene_id", "tx_name"))
# extract coding sequence regions

pseudo_gr <- gff[gff$type=="pseudogene"]
rrna_gr <- gff[gff$type=="rRNA"]
is_gr <- gff[grepl("insertion sequence", gff$mobile_element_type)]
gene_gr <- gff[gff$type=="gene"]

# filter gff files for pseudogenes, rrna, insertion sequences and genes

fa <- FaFile(fna_file)  
# creates a FaFile object for efficiently accessing the FASTA file
```
</div>

Then we work through each vcf file, extract the relevant information and annotate the mutations.

<div class="fold s">
```{r process_vcf, eval=process_vcf, message=FALSE, warning=FALSE}

vcf_files<-list.files(file.path(root, "vcf"))
vcf_files<-vcf_files[grep(".flt.vcf.gz$", vcf_files)]
# get a list filtered vcf file names 

sample<-unlist(lapply(strsplit(vcf_files, "_|\\."), function(x){x[2]}))
sample<-gsub("391-Ancestor", "391-Ancestor_1", sample)
sample<-gsub("392-Ancestor", "392-Ancestor_2", sample)
# get the sample names from the vcf file names. 

line<-sample_line$Sample.name[match(unlist(lapply(strsplit(sample, "-"), function(x){x[1]})), sample_line$Sample.number)]
# get the MA line names associated with each sample. 

vcf_dt<-c()

for(i in 1:length(vcf_files)){

	file.i<-file.path(root, "vcf", vcf_files[i])
	vcf <- readVcf(file.i)
	# read in the filtered vcf file

	vr  <- granges(vcf) 												# get the coordinates of the variants

	is_cds<-countOverlaps(vr, cds_gr)>0
  is_pseudo<-countOverlaps(vr, pseudo_gr)>0
  is_rrna<-countOverlaps(vr, rrna_gr)>0
  is_is<-countOverlaps(vr, is_gr)>0

  # evaluate whether the variants overlap coding sequence, 
  # pseudogenes, rrna genes and insertion sequences.

	coding <- predictCoding(vcf, txdb, seqSource = fa)
  # gets coding mutation types (synonymous, non-synonymous, )

	mini_vcf<-fread(file.i, sep = "\t", header = TRUE, skip = "#CHROM")
  # read vcf into a data.table

	n_ref<-as.vector(sapply(mini_vcf$INFO, function(x){sum(as.numeric(strsplit(strsplit(x, "DP4=|;MQ=")[[1]][2], ",")[[1]])[1:2])}))
	n_alt<-as.vector(sapply(mini_vcf$INFO, function(x){sum(as.numeric(strsplit(strsplit(x, "DP4=|;MQ=")[[1]][2], ",")[[1]])[3:4])}))
  # get number of reads in support of the reference and alternate allele

	is_nonsyn<-gene<-locus_tag<-rep(NA, length(names(vcf)))
	# empty vector for writing mutation type and gene name for mutations in coding sequence

	is_nonsyn[match(names(coding), names(vcf))]<-as.character(coding$CONSEQUENCE)
  # writes mutation type for mutations in coding sequence

  hits    <- findOverlaps(vr, gene_gr)
  # finds overlap between variant coordinates and genes

  gene[queryHits(hits)] <- unlist(gene_gr$Name[subjectHits(hits)])
  locus_tag[queryHits(hits)] <- unlist(gene_gr$locus_tag[subjectHits(hits)])
  # writes gene name for mutations falling in genes 

	vcf_dt <- rbind(vcf_dt, cbind(sample[i], line[i], is_cds, is_nonsyn, is_pseudo, is_rrna, is_is, gene, locus_tag, mini_vcf$QUAL, n_ref, n_alt, mini_vcf[,c(2,4,5)]))

  # consolidate information into a data.frame 
}

colnames(vcf_dt)[1:12]<-c("sample", "line", "coding", "type", "pseudo", "rrna", "is", "gene", "locus_tag", "QUAL", "n_ref", "n_alt")
```
</div>

# Filtering Mutations

`vcf_dt` contains `r nrow(vcf_dt)` variant calls that have not been filtered in any way. We first filter out the two variants that are present in the ancestor and therefore all MA lines. We then filter out sites of low quality to give us a set of possible mutations. These possible mutations are split in to a high confidence set (more than 30 reads and more than 95% of reads supporting the variant) and a low confidence set. The low confidence set are manually validated, and information is given as to why we think they are valid or not.   

## Ancestral mutations

Every line had an identical variant called at positions 4296380 and 4381253. These are mutations that have occurred in our ancestral line before MA commenced and so we remove them:

```{r }
orig_mut<-c(4296380, 4381253)

vcf_dt<-subset(vcf_dt, !POS%in%orig_mut)
```

## Variant calls with low Phred-scaled score (QUAL$<20$)

For the remaining `r nrow(vcf_dt)` variant calls, we can plot the proportion of reads that support the alternate allele against the number of reads that support the alternate allele. In black we have INDELS and in red we have bps. Triangles are mutations in rRNA genes or insertion sequences, and circles are the remainder.

```{r plot_all}
vcf_dt$prop<-vcf_dt$n_alt/(vcf_dt$n_ref+vcf_dt$n_alt)
plot(jitter(prop, amount=0.01)~jitter(n_alt,5), col=1+as.numeric(nchar(ALT)==nchar(REF)), cex=0.5, ylab=c("Proportion alternate"),  pch=c(1,17)[1+as.numeric(rrna | is)], xlab=c("Number alternate"), data=vcf_dt)
```

Note that most variant calls only have a single read in support of the alternate allele:

```{r }
table(vcf_dt$n_alt==1)
```

Of those that only have a single read in support of the alternate allele, most are either in rRNA or insertion sequences

```{r }
table(subset(vcf_dt, n_alt==1)$rrna | subset(vcf_dt, n_alt==1)$is)
```

If we exclude sites with low quality, most of these sites disappear and we are left with `r nrow(subset(vcf_dt, QUAL>20))` possible mutations:

```{r }
vcf_dt_qual<-subset(vcf_dt, QUAL>20)
```

## Possible mutations

For variants that pass the quality filter, we can see that bps generally have close to 100% of reads supporting the alternate, whereas for INDELS the proportion is generally less. 


```{r plot_qual}
plot(jitter(prop, amount=0.01)~jitter(n_alt,2), col=1+as.numeric(nchar(ALT)==nchar(REF)), cex=1, ylab=c("Proportion alternate"),  pch=c(1,17)[1+as.numeric(rrna | is)], xlab=c("Number alternate"), data=vcf_dt_qual)
abline(v=30, lty=2)
abline(h=0.95, lty=2)
```
with at least 30 reads supporting the alternate allele and less than 5% of reads supporting the reference allele

All mutations with with at least 30 reads in support of the alternate allele and more than 95% of reads in support of the alternate allele (i.e. the upper right-hand quadrant formed by the dotted lines) were considered unambiguously real. The remainder were checked in the Integrative Genomics Viewer [@Robinson.2011]. The `check` column of manual\_check.csv indicates whether the mutation was checked (not `NA`) and whether it was deemed OK (`OK`) or not (`NO`). 

```{r }
manual_check<-read.csv(file.path(root, "meta/manual_check.csv"))

pos_in_man<-match(paste(vcf_dt_qual$sample,vcf_dt_qual$POS), paste(manual_check$sample,manual_check$POS))

vcf_dt_qual$check<-manual_check$check[pos_in_man]
```

Note that 271-96 1196220, which met the strict filter, was also checked as it was in *icd* and had exactly 30 reads in support (see below)

## BPS with less than 95% support

We can look at the 6 bps with 95% of read support or less: 

```{r }
subset(vcf_dt_qual[,-c(2:5)], prop<=0.95 & nchar(ALT)==nchar(REF))
```

Two looked like mapping error (one is in a rRNA gene) although one of those deemed as an error (230-42 960164 T->G) could have been real if there was a lot of contamination - hard to verify since the line didn't have other mutations. Four looked OK,  but 202-10 226642 T->G is in an rRNA gene and so we removed it

```{r }
vcf_dt_qual$check[which(vcf_dt_qual$sample=="202-10" & vcf_dt_qual$POS=="226642")]<-"NO"
```

The two 279-107 mutations looked OK and there was perhaps some contamination as all mutations had some reads supporting the reference 

```{r }
subset(vcf_dt_qual[,-c(2:5)], sample=="279-107")
```

287-116 37246 looked OK - perhaps some contamination also.

## INDELS with less than 95% support

28 INDELS have less than 95% of read support. None occurred in rRNA genes or insertion sequences. 

```{r }
subset(vcf_dt_qual[,-c(2:5)], prop<=0.95 & nchar(ALT)!=nchar(REF))
```

The mutations

204-12  1198453 A->AATGAAATG  

262-84  1198436 A->AAATGAAAT 

312-145 1198453 AATGAAATG->A 

232-44  1198453 AATGAAATG->A 

were all deemed mapping errors as they were in same region across lines. Similarly, 

272-97  2765411 C->CGCACTATG 

298-128 2765411 CGCACTATG->C 

390-38  2765411 C->CGCACTATG

were deemed mapping errors for the same reason.

389-209 2560977 TGGTGGGTCAAAAGTTGCCGTTAA->T was deemed real but the mutation 389-209 2560999 A->AG unlikely to be real - only one mutation happened in this region.

The mutations

310-142 1624349 AT->A 

310-142 1892648 TACC->T

310-142 1986327 GA->G

310-142 4606086 AG->A

310-142 2185868 T->TG

were all deemed real, but sample 310-142 has 17 additional mutations, one in the mismatch repair gene *mutS*. There's no sign of contamination - it seems like these mutations occurred after the conclusion of mutation accumulation but during the growth stage prior to sequencing. 

The remaining mutations were all deemed correct. In most cases the INDEL fell in a repetitive region and the reads supporting the reference did not span sufficient repeats to call the mutation.

## BPS with fewer than 30 supporting reads 


We can also check the 24 bps with fewer than 30 reads in support of the alternate, yet the proportion exceeded 95%:

```{r }
subset(vcf_dt_qual[,-c(2:5)], n_alt<30 & prop>0.95 & nchar(ALT)==nchar(REF))
```

Four are in an rRNA gene and four are in an insertion sequence. Nine mutations are in the same line in the *icd* gene and were considered mapping errors (see below).  Of the remaining six, four looked real (minimum read support is 6) and two did not (read supports of one and two)

## INDELS with fewer than 30 supporting reads 

Three INDELS had fewer than 30 supporting reads but a proportion in excess of 0.95. 

```{r }
subset(vcf_dt_qual[,-c(2:5)], n_alt<30 & prop>0.95 & nchar(ALT)!=nchar(REF))
```

252-73 1386909 G->GC was deemed not real (only 2 reads with support) as it was adjoining a big region with no reads mapping - which was identified as a big deletion by Parliment.

365-200 4168938 GA->G was deemed not real (6 reads with support)

263-85 3946632 A->AT was deemed probably real (8 reads with support) and mapping was likely difficult due to a bps mutation at 3946620.


## rRNA and insertion sequence

In addition there is one bps mutation in an rRNA gene and an insertion sequence. Both have good support and are deemed real.

```{r }
subset(vcf_dt_qual[,-c(2:5)], n_alt>30 & prop>0.95 & (rrna | is))
```

## Large structural variants

Three large deletions were also found using Parliament2 [@Zarate.2021]: a 321bp deletion in sample 252-73 (1386588-1386909), a 15112bp deletion in sample 271-96 (1196300-1211412) and a 456bp deletion in sample 338-172 (2817640-2818096). All were checked in IGV.  Note that the deletion in sample 271-96 spans *icd* and the false mutations called were around the ends of this deletion. The deletion coincides with the cryptic prophage e14, which has previously been shown to excise on induction of the SOS pathway [@Brody.1985].

## Real mutations

We only retain those mutations that passed our criteria, and excluded the mutations in the hyper-mutator that arose after mutation accumulation 

```{r }
mutations_real<-subset(vcf_dt_qual, is.na(check) | check=="OK" & !(sample=="310-142" & POS%in%c(1624349,1892648,1986327,4606086,2185868)))
```

Note that if we had simply discarded all mutations in insertion sequence, rRNA genes and the gene *icd* we would have had a reasonable agreement: 

```{r }
with(vcf_dt_qual, table(problem=(is | rrna | gene=="icd"), pass=(is.na(check) | check=="OK")))
```

We add the large deletions manually to the list of real mutations:

```{r }
tmp<-matrix(NA, 3, ncol(mutations_real))
colnames(tmp)<-colnames(mutations_real)
tmp[,"sample"]<-c("252-73", "271-96", "338-172")
tmp[,"line"]<-c(115,42,215)
tmp[,"POS"]<-c(1386588, 1196300, 2817640)
tmp[,"ALT"]<-rep("<DEL>",3)
mutations_real<-rbind(mutations_real, tmp)
```
and the save the files

```{r eval=save}
write.csv(vcf_dt_qual, file.path(root, "meta/pos_mutations.csv"))
write.csv(mutations_real, file.path(root, "meta/real_mutations.csv"))
```

# False-negative rate for mutation calling

To determine how many mutations our pipeline may have missed, we took the reference sequence and randomly changed a base, added a base and deleted a base. We then mapped reads from one of our lines to this modified reference and called variants. We repeated this process for each of the 196 libraries and assessed how often the simulated mutations were called.  

## Simulate mutations

<div class="fold s">
```{r  message=FALSE, warning=FALSE, eval=sim_mutations}
library(stringi)

bases<-c("A", "C", "T", "G")

ref_fna<-scan(file.path(root, "ref/GCF_000005845.2_ASM584v2_genomic.fna"), what="character", sep="\n")
# read in original reference genome

header<-ref_fna[1]
ref_fna<-paste(ref_fna[-1], collapse="")

genome_size<-nchar(ref_fna)

file_names<-list.files(file.path(root, "raw"))
# fastq file names

mu_site<-matrix(sample(1:genome_size, size=length(file_names)*3), length(file_names),3)
# determine the position on which mutations happen 

test_mutations<-as.data.frame(matrix(NA, length(file_names)*3,10))
colnames(test_mutations)<-c("sample", "position", "ref", "alt", "n_ref", "n_alt", "REF", "ALT", "QUAL", "POS") 
# data frame for writing results

test_mutations[,"sample"]<-rep(file_names, each=3)

system(paste0("mkdir ", root, "/ref_test"))
# make a directory for writing the modified reference genomes

for(i in 1:length(file_names)){

   mu_orig<-rep(NA,3)

   new_fna<-ref_fna
   # duplicate original reference sequence

   for(j in 1:3){  # 1 is bps 2 is deletion 3 is insertion

     mu_orig[j]<-substr(ref_fna, mu_site[i,j],mu_site[i,j])
     # original base

     if(j==1){  # SNPS      
        test_mutations[(i-1)*3+j,"alt"]<-mu_orig[j]
        # original base is then identified as mutated base from the sequencing

        test_mutations[(i-1)*3+j,"ref"]<-sample(setdiff(bases,mu_orig[j]), size=1)
        stri_sub(new_fna, mu_site[i,j],mu_site[i,j])<-test_mutations[(i-1)*3+j,"ref"]
        # replace ref sample base with a new base which will be the reference allele

        test_mutations[(i-1)*3+j,"position"]<-mu_site[i,j]
     }

     if(j==2){ # insert into reference to make DELETIONS


        test_mutations[(i-1)*3+j,"ref"]<-sample(bases, size=1)
        # base to insert

        run<-0
        while(substr(ref_fna, mu_site[i,j]-run,mu_site[i,j]-run)==test_mutations[(i-1)*3+j,"ref"]){run<-run+1}
        # indicate the number of bases prior to the mutation that have the same sequence

        mu_site[i,j]<-mu_site[i,j]-run
        # place the mutation at the start of the run as makes no difference and indexed correctly in vcf

        test_mutations[(i-1)*3+j,"alt"]<-substr(ref_fna, mu_site[i,j], mu_site[i,j])
        # a random base is inserted into the original sequence and so a deletion is identified from the sequencing

        stri_sub(new_fna, mu_site[i,j],mu_site[i,j])<-paste0(test_mutations[(i-1)*3+j,"alt"], test_mutations[(i-1)*3+j,"ref"])
        # add the random base to the ref sample base which will be the reference allele

        test_mutations[(i-1)*3+j,"ref"]<-substr(new_fna, mu_site[i,j],mu_site[i,j]+1)

        test_mutations[(i-1)*3+j,"position"]<-mu_site[i,j]

     }
     if(j==3){ # delete from reference to get INSERTIONS

        run<-0
        while(substr(ref_fna, mu_site[i,j]-run,mu_site[i,j]-run)==mu_orig[j]){run<-run+1}

        # indicate the number of bases prior to the mutation that have the same sequence

        mu_site[i,j]<-mu_site[i,j]-run
        # place the mutation at the start of the run as makes no difference and indexed correctly in vcf

        test_mutations[(i-1)*3+j,"ref"]<-substr(ref_fna,mu_site[i,j],mu_site[i,j])

        test_mutations[(i-1)*3+j,"alt"]<-paste0(test_mutations[(i-1)*3+j,"ref"], mu_orig[j])
        # a random base is deleted from the original sequence and so the original base is identified as an insertion from the sequencing

        if(mu_site[i,2]<mu_site[i,3]){
          adjust<-1  
        }else{
          adjust<-0
        }
        # if a random base was added to the reference before the removal have to increment the position by one

        stri_sub(new_fna, mu_site[i,j]+adjust,mu_site[i,j]+adjust+1)<-test_mutations[(i-1)*3+j,"ref"]
        # delete a random base from ref and so this site will not be present in the ref

        test_mutations[(i-1)*3+j,"position"]<-mu_site[i,j]
     }
  }
  if(mu_site[i,3]<mu_site[i,2]){
    test_mutations[(i-1)*3+2,"position"]<-test_mutations[(i-1)*3+2,"position"]-1
  }
  if(mu_site[i,3]<mu_site[i,1]){
    test_mutations[(i-1)*3+1,"position"]<-test_mutations[(i-1)*3+1,"position"]-1
  }
  if(mu_site[i,2]<mu_site[i,3]){
    test_mutations[(i-1)*3+3,"position"]<-test_mutations[(i-1)*3+3,"position"]+1
  } 
  if(mu_site[i,2]<mu_site[i,1]){
    test_mutations[(i-1)*3+1,"position"]<-test_mutations[(i-1)*3+1,"position"]+1
  } 
   
  fna_path<-file.path(root, "ref_test", paste0(file_names[i], ".fna"))

  cat(header, file=fna_path, sep="\n")
  cat(new_fna, file=fna_path, sep="\n", append=TRUE)
  # write new reference to a file
}
```
</div>


## Call simulated mutations

For each library we can map the reads to the new reference genome and call mutations. To achieve this we use the scripts `map_test.sh` and `call_test.sh` which are minor modifications to the scripts `map.sh` and `call.sh`described above. The information is organised and written to test$_$mutations.csv where the columns `ref` and `alt` refer to the simulated mutation and `REF` and `ALT` refer to the mutation called. If a mutation wasn't called at the correct position `REF` and `ALT` are `NA`.

<div class="fold s">
```{r  message=FALSE, warning=FALSE, eval=sim_mutations}
system(file.path(root, "scripts/map_test.sh"))
# map reads to the respective reference

system(file.path(root, "scripts/call_test.sh"))
# call variants with respect to the respective reference

for(i in 1:length(file_names)){

  vcf_path<-file.path(root, "vcf_test", paste0(file_names[i], ".flt.vcf.gz"))
  mini_vcf<-fread(vcf_path, sep = "\t", header = TRUE, skip = "#CHROM")
  # read in vcf

  n_ref<-as.vector(sapply(mini_vcf$INFO, function(x){sum(as.numeric(strsplit(strsplit(x, "DP4=|;MQ=")[[1]][2], ",")[[1]])[1:2])}))
  n_alt<-as.vector(sapply(mini_vcf$INFO, function(x){sum(as.numeric(strsplit(strsplit(x, "DP4=|;MQ=")[[1]][2], ",")[[1]])[3:4])}))

  # number of reads in support of the reference and alternate

  qual<-mini_vcf$QUAL
  # quality score of the call

  pos<-mini_vcf$POS
  # vcf position

  ref<-mini_vcf$REF
  alt<-mini_vcf$ALT
  # called base

  hit<-sapply(test_mutations[(i-1)*3+1:3,"position"], function(x){dist<-abs(x-mini_vcf$POS); if(any(dist<6)){which.min(dist)}else{NA}})
  # find where the known mutations are in the vcf file (if they exist) - match to within 5bp in case there are position shifts
  
  if(any(!is.na(hit))){
    test_mutations[(i-1)*3+1:3,"n_ref"]<-n_ref[hit]
    test_mutations[(i-1)*3+1:3,"n_alt"]<-n_alt[hit]
    test_mutations[(i-1)*3+1:3,"REF"]<-ref[hit]
    test_mutations[(i-1)*3+1:3,"ALT"]<-alt[hit]
    test_mutations[(i-1)*3+1:3,"QUAL"]<-qual[hit]
    test_mutations[(i-1)*3+1:3,"POS"]<-pos[hit]
  }
}

system(paste("rm -r", file.path(root, "vcf_test")))
system(paste("rm -r", file.path(root, "bam_test")))
system(paste("rm -r", file.path(root, "filtered_bam_test")))
system(paste("rm -r", file.path(root, "ref_test")))
```
</div>

We can then assess how many mutations are called and whether they pass the strict filter

```{r message=FALSE, warning=FALSE, echo=FALSE}
if(save){
  write.csv(test_mutations, file.path(root, "meta/test_mutations.csv"))
}else{
  test_mutations<-read.csv(file.path(root, "meta/test_mutations.csv"))
}
```

```{r  message=FALSE, warning=FALSE}
test_mutations$strict<-with(test_mutations, n_alt>=30 & (n_alt/(n_alt+n_ref))>0.95 & QUAL>=20)
# determine those mutations that would pass the strict filter

test_mutations$mtype<-with(test_mutations, c("d","b","i")[nchar(alt)-nchar(ref)+2])

called<-with(test_mutations, table(mtype=mtype, called=!is.na(test_mutations$strict)))
# create a table of called vs. not called across mutation types

called

strict<-with(test_mutations, table(mtype=mtype, strict=strict))
# create a table of those called mutation passing vs. not the strict filter across mutation types

strict
```

Almost all mutations are called
